\chapter{XRM: A Conceptual Model for XR}
\label{ch:conceptual-model}

According to the bibliographical research in \autoref{sec:background-conceptual}, most of the models described are referred to the design of VEs, indeed there are few works dedicated to the design of mixed experiences. This means that existing models are created to meet the needs of a specific application, as the authors continue to build new models from scratch, which are closely tailored to their own domain and therefore difficult to adapt to other applications. The conceptual model should guide designers in the creation and design of multi-reality and multi-platform experiences in order to build a communication bridge between them and the team that has to implement the application. Often these are complex systems combining 3D elements, physical objects, "behaviour-rich content" \cite{walczak_structured_2008}, devices of different nature, single or shared experiences, indoor or outdoor experiences etc. this complexity severely challenge the developers, especially if they are always compelled to intervene directly on the code. 

The XRM Conceptual Model stems from the need to provide a tool capable of covering most of all the technologies included in the Virtual Continuum. The approach used is centered on the human, who is the protagonist in many scenarios thanks to the support of devices and platforms dependent on the environment experienced.. The aim is to create a means to enable high-level development of XR applications, in order to involve also non-experts in the domain. 
In the light of the previous comparative and illustrative analysis, from the variety of conceptual models and based on the planned features, the ISS Model was chosen as a starting point for the upcoming model. The model of Gianotti et al.~\cite{dobbie_modeling_2020} provided a dual structure that allows for the systematic design of the static and dynamic parts of an XR experience. Although the context considered have few points in common, the translation of the concepts was not too cumbersome and forced. The XRM Conceptual Model is divided into three parts: a Structural, a Behavioural and an Interaction sub-model, which are described in the next section. 
The Model has been added to the dual structure proposed by the conceptual models analysed in the Comparative study (\autoref{sec:background-conceptual}) with the intention of detailing the concepts separately before combining them. Initially, the elements will be presented by explaining what they represent, their external characteristics and initial properties. Then, user behaviours and effects on technological components will be defined. After that, all the previously defined concepts will be used to compose more or less complex activities that constitute the flow of the experience.


\input{Chapters/3.ConceptualModel/StructuralModel}
\input{Chapters/3.ConceptualModel/BehavioralModel}
\input{Chapters/3.ConceptualModel/InteractionModel}
\input{Chapters/3.ConceptualModel/NURE_Example}